FLM for growth as function of SPEI

### 2020-11-12 analysis/paper/vital rate models.Rmd
updated to work with new SPEI history matrix-dataframes

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(mgcv)
library(dlnm)
library(bbmle)
library(lubridate)
library(glue)
library(patchwork)
library(sf)

@@

# library(HeliconiaDemography) #load functions in this compendium
load_all()
```

@@

```{r data, echo=TRUE}
spei <- read_csv(here("analysis", "data", "derived_data", "spei_lags.csv"))
xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
demog <- read_rds(here("analysis", "data", "derived_data", "ha_survey.rds"))
plot_coords <- read_csv(here("analysis", "data", "raw_data", "plot_coords.csv"))
```
# Join data sets

@@

```{r}
xa2 <- 
  xa %>%
  filter(month(yearmonth) == 2) %>% 
  mutate(year = year(yearmonth), .after =yearmonth) %>% 
  separate(latlon, into = c("lat", "lon"), sep = "_")

```

@@

```{r}
unique(demog$plot)
unique(demog$bdffp_reserve_no)
unique(spei$site)

demog %>% filter(code_notes ==  "dead (2)", year == 2009)
```

@@

```{r}
demog_full <-
  left_join(demog, spei, by = c("rain_site" = "site", "year" = "census_year")) %>% 
  mutate(log_shts = log(shts), .after = shts)
```

@@

```{r}
Q_post_cf <-
  demog_post_cf %>% 
  select(starts_with("spei_")) %>% 
  as.matrix()

Q_post_1ha <-
  demog_post_1ha %>% 
  select(starts_with("spei_")) %>% 
  as.matrix()

Q_sdlg_cf <-
  demog_sdlg_cf %>% 
  select(starts_with("spei_")) %>% 
  as.matrix()

Q_sdlg_1ha <-
  demog_sdlg_1ha %>% 
  select(starts_with("spei_")) %>% 
  as.matrix()

L <-
  matrix(1:24,
         nrow = nrow(demog_full),
         ncol = 24,
         byrow = TRUE)

colnames(L) <- paste0("L_", 1:24)

L_post_cf <- L[1:nrow(Q_post_cf), ]
L_post_1ha <- L[1:nrow(Q_post_1ha), ]
L_sdlg_cf <- L[1:nrow(Q_sdlg_cf), ]
L_sdlg_1ha <- L[1:nrow(Q_sdlg_1ha), ]
```

@@

```{r}
g_post_cf <-
  gam(shts_next ~ 
        s(log_shts) + #does this need to be log # shoots?
        # s(shts) +
        s(Q_post_cf, L_post_cf,
          bs = "cb",
          k = c(3, 24),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "log"),
      data = demog_post_cf,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

@@

```{r}
g_post_1ha <-
  gam(shts_next ~ 
        s(log_shts) +
        s(Q_post_1ha, L_post_1ha,
          bs = "cb",
          k = c(3, 24),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "log"),
      data = demog_post_1ha,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

@@

```{r}
pred_g_post_cf <- cb_margeff(Q_post_cf, L_post_cf, g_post_cf, ref_data = ref_data)
pred_g_post_1ha <- cb_margeff(Q_post_1ha, L_post_1ha, g_post_1ha, ref_data = ref_data)
```

@@

```{r}
# spei_1 is january
labs <- c("Jan", paste0(month(12:1, label = TRUE), " (y-1)"), paste0(month(12:2, label = TRUE), " (y-2)"))
```

@@

```{r}
p_growth_cf <-
  pred_g_post_cf %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 1:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
  labs(title = "Continuous Forest") +
  theme_bw()
# p_growth_cf
```

```{r}
p_growth_1ha <-
  pred_g_post_1ha %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 1:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
  labs(title = "1-ha Fragments") +
  theme_bw()
# p_growth_1ha
```

@@

```{r}
p_growth_cf_slice <-
  pred_g_post_cf %>% 
  filter(nearest(x, -1.50)) %>% 
  ggplot(aes(x = lag, y = exp(fitted))) +
  geom_ribbon(aes(
    ymin = exp(fitted - se.fit * 1.96),
    ymax = exp(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 1:24, labels = labs) +
  labs(title = "Continuous Forest", y = "shoots (t+1)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_growth_cf_slice
```

```{r}
p_growth_1ha_slice <-
  pred_g_post_1ha %>% 
  filter(nearest(x, -1.5)) %>% 
  ggplot(aes(x = lag, y = exp(fitted))) +
  geom_ribbon(aes(
    ymin = exp(fitted - se.fit * 1.96),
    ymax = exp(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 1:24, labels = labs) +
  labs(title = "1-ha Fragments", y = "shoots (t+1)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_growth_1ha_slice
```

@@

foo <- function(x) {
  d <- 1:24
  names(d)<-labs
  x <- as.numeric(x)
  # s <- glue("lag {x} ({names(d[x])})")
  s <- names(d[x])
  return(s)
}

@@

```{r}
s_post_cf <-
  gam(surv ~ 
        s(log_shts) +
        s(Q_post_cf, L_post_cf,
          bs = "cb",
          k = c(3, 24),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
anova(s_post_cf)
```

```{r}
s_post_1ha <-
  gam(surv ~ 
        s(log_shts) +
        s(Q_post_1ha, L_post_1ha,
          bs = "cb",
          k = c(3, 24),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
anova(s_post_1ha)
```

@@

```{r}
#big plants
# ref_data <- tibble(log_shts = 1.6)

pred_s_post_cf <- cb_margeff(Q_post_cf, L_post_cf, s_post_cf, ref_data = ref_data)
pred_s_post_1ha <- cb_margeff(Q_post_1ha, L_post_1ha, s_post_1ha, ref_data = ref_data)
```

@@

```{r}
p_surv_cf <-
  pred_s_post_cf %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 1:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
  labs(title = "Continuous Forest") +
  theme_bw()
p_surv_cf
```

```{r}
p_surv_1ha <-
  pred_s_post_1ha %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 1:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
  labs(title = "1-ha Fragments") +
  theme_bw()
p_surv_1ha
```

@@

```{r}
p_surv_cf_slice <-
  pred_s_post_cf %>% 
  filter(nearest(x, -1.50)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 1:24, labels = labs) +
  labs(title = "Continuous Forest", y = "P(survival)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_surv_cf_slice
```

```{r}
p_surv_1ha_slice <-
  pred_s_post_1ha %>% 
  filter(nearest(x, -1.5)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 1:24, labels = labs) +
  labs(title = "1-ha Fragments", y = "P(survival)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_surv_1ha_slice
```

@@

```{r include=FALSE}
g_cf <-  plot.gam(g_post_cf,  shift = coef(g_post_cf)[1])
g_1ha <- plot.gam(g_post_1ha, shift = coef(g_post_1ha)[1])

@@

It looks like drought in October is bad for growth in both CF and fragments, but drought in June and July is much worse for plants in continuous forest.

Reminder that these are still fit with two different models and that `predict()` is using different climate histories (`Q`).  Might not be entirely comparable?

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print=FALSE)
library(tidyverse)
library(here)
library(mgcv)
library(dlnm)
library(bbmle)
library(lubridate)
library(glue)
library(patchwork)
library(sf)
library(tsibble)

@@

# library(HeliconiaDemography) #load functions in this compendium
# devtools::load_all()
```

@@

```{r data, echo=TRUE}
spei <- read_rds(here("analysis", "data", "derived_data", "spei_lags.rds"))
xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
demog <- read_rds(here("analysis", "data", "derived_data", "ha_survey.rds"))
plot_coords <- read_csv(here("analysis", "data", "raw_data", "plot_coords.csv"))
```
# Join data sets

@@

```{r}
xa2 <- 
  xa %>%
  filter(month(yearmonth) == 2) %>% 
  mutate(year = year(yearmonth), .after =yearmonth) %>% 
  separate(latlon, into = c("lat", "lon"), sep = "_")
xa2
```

@@

```{r}
unique(demog$plot)
unique(demog$bdffp_reserve_no)
unique(spei$site)
```

@@

Get SPEI history only for census months (assuming all censuses done in February)

```{r}
spei_census <- 
  spei %>% 
  filter(month(yearmonth) == 2) %>% 
  mutate(year = year(yearmonth), .before = site) %>% 
  select(-yearmonth)
```

Join

```{r}
demog_full <-
  left_join(demog, spei_census, by = c("rain_site" = "site", "year")) %>% 
  mutate(log_shts = log(shts), .after = shts)

# demog_full %>% select(spei, spei_history)
```

@@

@@

```{r}
g_post_cf <-
  gam(shts_next ~ 
        s(log_shts) + #does this *need* to be log # shoots?
        # s(shts) +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "log"),
      data = demog_post_cf,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

@@

```{r}
g_post_1ha <-
  gam(shts_next ~ 
        s(log_shts) +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "log"),
      data = demog_post_1ha,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

@@

```{r}
pred_g_post_cf  <- cb_margeff(g_post_cf, Q = spei_history, L = L, ref_data = ref_data)
pred_g_post_1ha <- cb_margeff(g_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
```

@@

```{r}
# first column is February, the month of the census
labs <- c("Feb", "Jan", paste0(month(12:1, label = TRUE), " (y-1)"), paste0(month(12:2, label = TRUE), " (y-2)"))
```

@@

```{r}
p_growth_cf <-
  pred_g_post_cf %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
  labs(title = "Continuous Forest") +
  theme_bw()
p_growth_cf
```

```{r}
p_growth_1ha <-
  pred_g_post_1ha %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
  labs(title = "1-ha Fragments") +
  theme_bw()
p_growth_1ha
```

@@

```{r}
p_growth_cf_slice <-
  pred_g_post_cf %>% 
  filter(nearest(x, -1.50)) %>% 
  ggplot(aes(x = lag, y = exp(fitted))) +
  geom_ribbon(aes(
    ymin = exp(fitted - se.fit * 1.96),
    ymax = exp(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 0:24, labels = labs) +
  labs(title = "Continuous Forest", y = "shoots (t+1)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_growth_cf_slice
```

```{r}
p_growth_1ha_slice <-
  pred_g_post_1ha %>% 
  filter(nearest(x, -1.5)) %>% 
  ggplot(aes(x = lag, y = exp(fitted))) +
  geom_ribbon(aes(
    ymin = exp(fitted - se.fit * 1.96),
    ymax = exp(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 0:24, labels = labs) +
  labs(title = "1-ha Fragments", y = "shoots (t+1)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_growth_1ha_slice
```

@@

foo <- function(x) {
  d <- 1:25
  names(d)<-labs
  x <- as.numeric(x)
  # s <- glue("lag {x} ({names(d[x])})")
  s <- names(d[x])
  return(s)
}

@@

```{r}
s_post_cf <-
  gam(surv ~ 
        s(log_shts) +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
anova(s_post_cf)
```

```{r}
s_post_1ha <-
  gam(surv ~ 
        s(log_shts) +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
anova(s_post_1ha)
```

@@

```{r}
pred_s_post_cf  <- cb_margeff(s_post_cf,  Q = spei_history, L = L, ref_data = ref_data)
pred_s_post_1ha <- cb_margeff(s_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
```

@@

```{r}
p_surv_cf <-
  pred_s_post_cf %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
  labs(title = "Continuous Forest") +
  theme_bw()
p_surv_cf
```

```{r}
p_surv_1ha <-
  pred_s_post_1ha %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
  labs(title = "1-ha Fragments") +
  theme_bw()
p_surv_1ha
```

@@

```{r}
p_surv_cf_slice <-
  pred_s_post_cf %>% 
  filter(nearest(x, -1.50)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 0:24, labels = labs) +
  labs(title = "Continuous Forest", y = "P(survival)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_surv_cf_slice
```

```{r}
p_surv_1ha_slice <-
  pred_s_post_1ha %>% 
  filter(nearest(x, -1.5)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 0:24, labels = labs) +
  labs(title = "1-ha Fragments", y = "P(survival)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_surv_1ha_slice
```

@@

```{r include=FALSE}
#TODO: switch to predict() so y axis is correct
g_cf <-  plot.gam(g_post_cf,  shift = coef(g_post_cf)[1])
g_1ha <- plot.gam(g_post_1ha, shift = coef(g_post_1ha)[1])

@@

It looks like drought in October is bad for growth in fragments, but drought in June and July is much worse for plants in continuous forest.

Reminder that these are still fit with two different models and that `predict()` is using different climate histories (`spei_history`).  Might not be entirely comparable?  Different penalties to the smooths make these hard to compare as well.
